\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-21 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add and begin to work on JSS-article-3

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validation of Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validation of Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validation of Regression Models"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples: See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

@@

Cross-validating mixed-effects models involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.

The \code{cvSelect()} function, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors.

@@

In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel. For example, the \code{cv()} and \code{cvSelect()} functions can perform computations in parallel and can independently replicate a cross-validation procedure several times. These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.

@@

```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5."
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=4)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=4,
       title="Degree", inset=0.02)
```

@@

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-selection}.

@@

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

@@

microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10
)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions.

@@

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter in a vignette for the \pkg{cv} package.

@@

The \code{cvSelect()} function in the \pkg{cv} package cross-validates a model-specification process in a general manner. Functions for three such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package [@FoxWeisberg:2019], transforms predictors and the response in a regression model towards normality; and \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially. In the vignette on extending the \pkg{cv} package, we explain how to add model-selection procedures.

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
```{r Auto-explore}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the numeric variables in the Auto data"
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3.4", "3.4", "5.6", "5.6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

@@

[^year]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process. It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation. We believe, however, that it's still useful to automate as much of the process as we can to obtain a more realistic, if still biased, estimate of the predictive power of a model.

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the working model fit to the Auto data"
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```
The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.

Some background: As @Weisberg:2014 [Sec. 8.2] explains, there are technical advantages to having (numeric) predictors in linear regression analysis that are themselves linearly related. If the predictors *aren't* linearly related, then the relationships between them can often be straightened by power transformations. Transformations can be selected after graphical examination of the data, or by analytic methods, such as transforming the predictors towards multivariate normality, which implies linearity. Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the conditional distribution of the response variable towards normality. Selecting transformations analytically raises the possibility of automating the process, as would be required for cross-validation.

@@

The transformed numeric variables are much better-behaved (cf., Figures \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and \ref{fig:Auto-transformed-scatterplot-matrix}):
```{r Auto-transformed-scatterplot-matrix}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the Auto data"
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
```
And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figures \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and \ref{fig:Auto-CR-plots-transformed}):
```{r Auto-CR-plots-transformed}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the model fit to the transformed Auto data"
crPlots(m, pch=".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log(n)$):
```{r}
library("MASS")
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
summary(m.step)
```
The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}. We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model:
```{r MSE-whole-selected-model}
mse(Auto$mpg, exp(fitted(m.step)))
```
This is considerably smaller than the MSE for the original working model:
```{r MSE-working-model}
mse(Auto$mpg, fitted(m.auto))
```
A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^medAbsErr]

@@

The \code{cvSelect()} function in the \pkg{cv} package allows us to cross-validate the whole model-selection procedure. The first argument to \code{cvSelect()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
```{r Auto-transform-and-select}
num.predictors
cvs <- cvSelect(selectTransStepAIC, data=Auto, seed=76692, model=m.auto,
                predictors=num.predictors,
                response="mpg", AIC=FALSE)
cvs

compareFolds(cvs)
```
The other arguments to \code{cvSelect()} are (see \code{?cvSelect} for additional optional arguments and details):

* \code{data}, the data set to which the model is fit.
* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
* Arguments required by the model-selection function: the starting \code{model} (\code{m.auto}) for predictor selection, the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation, and \code{AIC=FALSE}, which specifies use of the BIC for model selection.

@@

* \code{selectTransStepAIC()} automatically computes CV cost criteria, here the MSE, on the \emph{untransformed} response scale.
* The estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models. Recall that when we selected predictors for the full data, we obtained a model with \code{horsepower}, \code{weight}, \code{acceleration}, \code{year}, and \code{origin}.

@@

The \pkg{cv} package is designed to be extensible in several directions; in order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with \code{selectModel()}. In this section, we illustrate (1) and (2); more extensive examples may be found in the vignette on extending the \pkg{cv} package.

@@

The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}, which was the focus of the original research on which this example is based [@AndersenHeathSinnott:2002]; the other predictors enter the model additively:
```{r BEPS-model}
library("nnet", quietly=TRUE)
m.beps <- multinom(vote ~ age + gender + economic.cond.national +
                       economic.cond.household + Blair + Hague + Kennedy +
                       Europe*political.knowledge, data=BEPS, trace=FALSE)

```
We set the argument \code{trace=FALSE} in the call to \code{multinom()} to suppress reporting the iteration history, which will be particularly appreciated when we repeatedly refit the model in cross-validation.

The \code{Europe} $\times$ \code{political.knowledge} interaction is associated with a very small $p$-value. Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the interaction in a "stacked-area" graph:
```{r BEPS-plot, fig.width=9, fig.height=5}
#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study."
plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
            xlevels=list(Europe=1:11, political.knowledge=0:3),
            fixed.predictors=list(given.values=c(gendermale=0.5))),
     lines=list(col=c("blue", "red", "orange")),
     axes=list(x=list(rug=FALSE), y=list(style="stacked")))
```

To cross-validate this multinomial-logit model we need an appropriate cost criterion. None of the criteria in the \pkg{cv} package will do---for example, \code{mse()} is appropriate only for a numeric response. The \code{BayesRule()} criterion, also supplied by \pkg{cv}, which is for a binary response, comes close:
```{r BayesRule}
BayesRule
```
After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction and then reports the proportion of incorrect predictions. Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, making possible the computation of bias correction and confidence intervals (as discussion in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

@@

head(GetResponse(m.beps))
```
and to try again:
```{r BEPS-test-default-2, error=TRUE}
cv(m.beps, seed=3465, criterion=BayesRuleMulti)
```
A \code{traceback()} (not shown) reveals that the problem is that the default method of \code{cv()} calls the \code{"multinom"} method for \code{predict()} with the argument \code{type="response"}, when the correct argument should be \code{type="class"}.  We therefore must write a \code{"multinom"} method for \code{cv()}, but that proves to be very simple:
```{r cv.nultinom}
cv.multinom <- function (model, data, criterion=BayesRuleMulti, k, reps,
                         seed, ...){
  NextMethod(type="class", criterion=criterion)
}
```
That is, we simply invoke the default \code{cv()} method via \code{NextMethod()}, with the \code{type} argument properly set. In addition to supplying the correct \code{type} argument, our method sets the default \code{criterion} for the \code{cv.multinom()} method to \code{BayesRuleMulti}.

@@

>>>>

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validating Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validating Regression Models"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, the results can vary substantially based on the random division of the data, particularly in smaller samples. See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

@@

Cross-validating mixed-effects models, using the \code{"glmmTMB"}, \code{"lme"}, and  \code{"merMod"} methods for \code{cv()}, involves special considerations that we take up in Section \ref{cross-validating-mixed-effects-models}.

The \code{"function"} method for \code{cv()}, discussed in Section \ref{cross-validating-model-specification}, cross-validates a complex model-specification process that may, for example, involve choice of data transformations and predictors, or model selection via CV itself.

@@

A number of existing \proglang{R} packages include functions for cross-validating regression models. We briefly situate the \pkg{cv} package relative to other \proglang{R} software for cross-validation in Section \ref{comparing-cv-to-other-r-software-for-cross-validation}.

In the interest of brevity, we won't describe all of the features of the \pkg{cv} package here, concentrating on the aspects of the package that are relatively novel. For example, the \code{cv()} function can perform computations in parallel and can independently replicate a cross-validation procedure several times. These and other features not discussed in this paper are taken up in the vignettes distributed with the package, which also provides greater detail on some of topics that we do describe, such as extensions to the package.

@@

```{r mpg-horsepower-scatterplot-polynomials, echo=FALSE}
#| out.width = "50%",
#| fig.height = 5,
#| fig.cap = "\\code{mpg} vs. \\code{horsepower} for the \\code{Auto} data, showing fitted polynomials of degree 1 through 5."
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=4)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=4,
       title="Degree", inset=0.02)
```

@@

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.

@@

* \code{method}, the computational method to employ: \code{"hatvalues"} is relevant only for LOO CV and bases computation on the hatvalues for the linear model; \code{"Woodbury"} employs the Woodbury matrix identity to compute the CV criterion with each fold deleted; \code{"naive"} updates the model using the \code{update()} function; and \code{"auto"} selects \code{"hatvalues"} for LOO CV and \code{"Woodbury"} for $k$-fold CV, both of which are much more efficient than literally updating the least-squares fit (see below and Section \ref{efficient-computations-for-linear-and-generalized-linear-models}).

@@

microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10
)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.

<!-- [^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last column, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions. -->

@@

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package, for example, is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter in a vignette for the \pkg{cv} package.

@@

The \code{"function"} method for \code{cv()} cross-validates a model-specification process in a general manner. Functions for four such model-specification processes are included in the package: \code{selectStepAIC()}, based on the \code{stepAIC()} function in the \pkg{MASS} package [@VenablesRipley:2002], performs stepwise predictor selection for regression models; \code{selectTrans()}, based on the \code{powerTransform()} function in the \pkg{car} package [@FoxWeisberg:2019], transforms predictors and the response in a regression model towards normality; \code{selectTransAndStepAIC()}---the use of which we illustrate in the current section---performs both of these procedures sequentially; and \code{selectModelList()}---also illustrated in the current section---uses CV both to select one of several competing models, and then, recursively, to estimate prediction error for the selected model. In a vignette on extending the \pkg{cv} package, we explain how to add model-selection procedures.

## Example: Data transformation and predictor selection for the Auto data

@@

The \code{Auto} data appeared in a preliminary example in Section \ref{preliminary-example-polynomial-regression}, where we employed CV to inform the selection of the degree of a polynomial regression of \code{mpg} on \code{horsepower}. Here, we consider more generally the problem of predicting \code{mpg} from the other variables in the \code{Auto} data. We begin with a bit of data management, and then examine the pairwise relationships among the numeric variables in the data set (Figure \ref{fig:Auto-explore}, produced by the \code{scatterplotMatrix()} function in the \pkg{car} package):
```{r Auto-explore}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the numeric variables in the \\code{Auto} data"
Auto$cylinders <- factor(Auto$cylinders,
                         labels=c("3-4", "3-4", "5-6", "5-6", "8"))
Auto$year <- as.factor(Auto$year)
Auto$origin <- factor(Auto$origin,
                      labels=c("America", "Europe", "Japan"))
rownames(Auto) <- make.names(Auto$name, unique=TRUE)
Auto$name <- NULL

@@

[^year]: Making the decision to treat \code{year} as a factor on this basis could be construed as cheating in the current context, which illustrates the difficulty of automating the whole model-selection process. It's rarely desirable, in our opinion, to forgo exploration of the data to ensure the purity of model validation. We believe, however, that it's still useful to automate as much of the process as we can, as illustrated here, to obtain a more realistic, if still biased, estimate of the predictive power of a model.

We start with a "working model" that specifies linear partial relationships of the response to the numeric predictors:
```{r Auto-working-model}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the working model fit to the \\code{Auto} data"
m.auto <- lm(mpg ~ ., data = Auto)
crPlots(m.auto, pch=".")
```
The component+residual plots in Figure \ref{fig:Auto-working-model} clearly reveal the inadequacy of the model.

Some background: As @Weisberg:2014 [Sec. 8.2] explains, there are technical advantages to having (numeric) predictors in linear regression analysis that are themselves linearly related. If the predictors *aren't* linearly related, then the relationships between them can often be straightened by power transformations. Transformations can be selected after graphical examination of the data, or by analytic methods, such as transforming the predictors towards multivariate normality, which implies linearity. Once the relationships between the predictors are linearized, it can be advantageous similarly to transform the conditional distribution of the response variable towards normality. Selecting transformations analytically raises the possibility of automating the process, as required for cross-validation.

@@

The transformed numeric variables are much better-behaved (cf., Figures \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and \ref{fig:Auto-transformed-scatterplot-matrix}):
```{r Auto-transformed-scatterplot-matrix}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data"
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
```
And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
```{r Auto-CR-plots-transformed}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the model fit to the transformed \\code{Auto} data"
crPlots(m, pch=".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
```{r}
library("MASS")
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
summary(m.step)
```
The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}. We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model:
```{r MSE-whole-selected-model}
mse(Auto$mpg, exp(fitted(m.step)))
```
This is considerably smaller than the MSE for the original working model:
```{r MSE-working-model}
mse(Auto$mpg, fitted(m.auto))
```
A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^medAbsErr]

@@

The \code{"function"} method for \code{cv()} allows us to cross-validate the whole model-selection procedure. The first argument to \code{cv()}---here \code{selectTransAndStepAIC}---is a model-selection function capable of refitting the model with a fold omitted and returning a CV criterion:
```{r Auto-transform-and-select}
num.predictors
cvs <- cv(selectTransStepAIC, data=Auto, seed=76692, 
          working.model=m.auto,
          predictors=num.predictors,
          response="mpg", AIC=FALSE)
cvs

compareFolds(cvs)
```
The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):

* \code{data}, the data set to which the model is fit.
* \code{seed}, an optional seed for \proglang{R}'s pseudo-random-number generator; as for \code{cv()}, if the seed isn't supplied by the user, then a seed is randomly selected and saved.
* Arguments required by the model-selection function: the starting \code{working.model} (\code{m.auto}) for transformation and predictor selection, the names of the variables---\code{predictors} and \code{response}---that are candidates for transformation, and \code{AIC=FALSE}, which specifies use of the BIC for model selection.

@@

* \code{selectTransStepAIC()} automatically computes CV cost criteria, here the MSE, on the \emph{untransformed} response scale.
* As we anticipated, the estimate of the MSE that we obtain by cross-validating the whole model-specification process is larger than the MSE computed for the model we fit to the \code{Auto} data separately selecting transformations of the predictors and the response and then selecting predictors for the whole data set.
* When we look at the transformations and predictors selected with each of the 10 folds omitted (i.e., the output of \code{compareFolds()}), we see that there is little uncertainty in choosing variable transformations (the \code{lam.*}s for the $x$s and \code{lambda} for $y$ in the output), but considerably more uncertainty in subsequently selecting predictors: \code{horsepower}, \code{weight}, and \code{year} are always included among the selected predictors; \code{acceleration} and \code{displacement} are included respectively in 4 and 3 of 10 selected models; and \code{cylinders} and \code{origin} are each included in only 1 of 10 models. Recall that when we selected predictors for the full data, we obtained a model with \code{horsepower}, \code{weight}, \code{acceleration}, \code{year}, and \code{origin}.

## Example: Applying recursive CV to polynomial regression for the Auto data

@@

The \pkg{cv} package is designed to be extensible in several directions; in order of increasing general complexity, we can add: (1) a cross-validation cost criterion; (2) a model class that's not directly accommodated by the \code{cv()} default method or by another directly inherited method; and (3) a new model-selection procedure suitable for use with the \code{"function"} method for \code{cv()}. In this section, we illustrate (1) and (2); more extensive examples may be found in the vignette on extending the \pkg{cv} package.

@@

The model fit to the data includes an interaction between \code{Europe} and \code{political.knowledge}, which was the focus of the original research on which this example is based [@AndersenHeathSinnott:2002]; the other predictors enter the model additively:
```{r BEPS-model}
library("nnet", quietly=TRUE)
m.beps <- multinom(vote ~ age + gender + economic.cond.national +
                       economic.cond.household + Blair + Hague + Kennedy +
                       Europe*political.knowledge, data=BEPS)

```
<!-- We set the argument \code{trace=FALSE} in the call to \code{multinom()} to suppress reporting the iteration history, which will be particularly appreciated when we repeatedly refit the model in cross-validation.  -->

The \code{Europe} $\times$ \code{political.knowledge} interaction is associated with a very small $p$-value. Figure \ref{fig:BEPS-plot} shows an "effect plot," using the the \pkg{effects} package [@FoxWeisberg:2019] to visualize the interaction in a "stacked-area" graph:
```{r BEPS-plot, fig.width=9, fig.height=5}
#| fig.cap = "Effect plot for the interaction between attitude towards European integration and political knowledge in the multinomial logit model fit to voting data from the 2001 British Election Panel Study."
plot(effects::Effect(c("Europe", "political.knowledge"), m.beps,
            xlevels=list(Europe=1:11, political.knowledge=0:3),
            fixed.predictors=list(given.values=c(gendermale=0.5))),
     lines=list(col=c("blue", "red", "orange")),
     axes=list(x=list(rug=FALSE), y=list(style="stacked")))
```
As political knowledge increases, voters tend to align their votes more closely with the party positions on European integration: The Conservative Party was "Eurosceptic," while Labour and the Liberal Democrats were supportive of the UK's participation in the EU.

To cross-validate this multinomial-logit model we need an appropriate cost criterion. None of the criteria in the \pkg{cv} package will do---for example, \code{mse()} is appropriate only for a numeric response. The \code{BayesRule()} criterion, also supplied by \pkg{cv}, which is for a binary response, comes close:
```{r BayesRule}
BayesRule
```
After doing some error checking, \code{BayesRule()} rounds the predicted proabability of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction and then reports the proportion of incorrect predictions. Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussion in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

@@

head(GetResponse(m.beps))
```
and to try again:
```{r BEPS-test-default-2, error=TRUE}
cv(m.beps, seed=3465, criterion=BayesRuleMulti)
```
A \code{traceback()} (not shown) reveals that the problem is that the default method of \code{cv()} calls the \code{"multinom"} method for \code{predict()} with the argument \code{type="response"}, when the correct argument should be \code{type="class"}.  We therefore must write a \code{"multinom"} method for \code{cv()}, but that proves to be very simple:
```{r cv.nultinom}
cv.multinom <- function (model, data, criterion = BayesRuleMulti, 
                         k, reps, seed, ...) {
    model <- update(model, trace = FALSE)
    NextMethod(
      type = "class",
      criterion = criterion,
      criterion.name = deparse(substitute(criterion))
    )
  }
```
That is, we simply call the default `cv()` method with the `type` argument properly set. In addition to supplying the correct `type` argument, our method sets the default `criterion` for the `cv.multinom()` method to `BayesRuleMulti`. Adding the argument `criterion.name=` `deparse(substitute(criterion))` is inessential, but it insures that printed output will include the name of the criterion function that's employed, whether it's the default `BayesRuleMulti` or something else. Prior to invoking `NextMethod()`, we called `update()` with `trace=FALSE` to suppress the iteration history reported by default by `multinom()`---it would be tedious to see the iteration history for each fold.

@@

# Comparing cv to other R software for cross-validation
